---
title: "FML_Assignment2_KNN"
author: "Mohana Krishna Nitesh Varma Kalidindi"
date: "2023-10-01"
output: html_document
---
# PROBLEM STATEMENT
Universal bank is a young bank growing rapidly in terms of overall customer acquisition.
The majority of these customers are liability customers (depositors) with varying sizes of
relationship with the bank. The customer base of asset customers (borrowers) is quite
small, and the bank is interested in expanding this base rapidly in more loan business. In
particular, it wants to explore ways of converting its liability customers to personal loan
customers.
A campaign that the bank ran last year for liability customers showed a healthy conversion
rate of over 9% success. This has encouraged the retail marketing department to devise
smarter campaigns with better target marketing. The goal is to use k-NN to predict whether
a new customer will accept a loan offer. This will serve as the basis for the design of a new
campaign.
The file UniversalBank.csv contains data on 5000 customers. The data include customer
demographic information (age, income, etc.), the customer’s relationship with the bank
(mortgage, securities account, etc.), and the customer response to the last personal loan
campaign (Personal Loan). Among these 5000 customers, only 480 (= 9.6%) accepted the
personal loan that was offered to them in the earlier campaign.

# SOLUTION

```{r}
#Read the csv file 
myUniversalData <- read.csv("UniversalBank.csv")

#Summary of the data
summary(myUniversalData)

#Structure of the data
str(myUniversalData)

#Transpose of the data frame
t(t(names(myUniversalData)))

# Loading all the required libraries
library(caret)
library(class)
library(e1071)
library(knitr)

# Drop ID and ZIP code from the dataset
myUniversalData <- myUniversalData[,-c(1,5)]

# Transform categorical variables into dummy variables

# Converting Education to a factor
myUniversalData$Education <- as.factor(myUniversalData$Education)

# Now, convert Education to Dummy Variables
groups <- dummyVars(~., data = myUniversalData) # This creates the dummy groups

universal_m.df <- as.data.frame(predict(groups,myUniversalData))

# Split the data to 60% training and 40 % Validation 

set.seed(1)  
train.index <- sample(row.names(universal_m.df), 0.6*dim(universal_m.df)[1])
valid.index <- setdiff(row.names(universal_m.df), train.index)  
train.df <- universal_m.df[train.index,]
valid.df <- universal_m.df[valid.index,]
t(t(names(train.df)))

# Normalize the data

train.norm.df <- train.df[,-10] # Personal Income is the 10th variable.
valid.norm.df <- valid.df[,-10]

norm.values <- preProcess(train.df[, -10], method=c("center", "scale"))
train.norm.df <- predict(norm.values, train.df[, -10])
valid.norm.df <- predict(norm.values, valid.df[, -10])
```

### Question - 1
Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, Securities Account = 0, CD Account = 0, Online = 1, and Credit Card = 1. Perform a k-NN classification with all predictors except ID and ZIP code using k = 1. Remember to transform categorical predictors with more than two categories into dummy variables first. Specify the success class as 1 (loan acceptance), and use the default cutoff value of 0.5. How would this customer be classified?

```{r}
# We have already converted all categorical variables to dummy variables
# Classify the given customer
given_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1)

# Normalize the given customer data
given.cust.norm <- given_customer
given.cust.norm <- predict(norm.values, given.cust.norm)

#  Classify the customer using k-NN with k=1
knn_result <- class::knn(train = train.norm.df, 
                       test = given.cust.norm, 
                       cl = train.df$Personal.Loan, k = 1)
knn_result

```

#### Determine how the customer would be classified
```{r}
if (knn_result == 1) {
  cat("The customer is classified as likely to accept the loan offer (Loan Acceptance = 1).\n")
} else {
  cat("The customer is classified as unlikely to accept the loan offer (Loan Acceptance = 0).\n")
}
```

### Question - 2
What is a choice of k that balances between overfitting and ignoring the predictor information?

#### The best choice of k which also balances the model from overfitting is k = 3
```{r}

# Calculate the accuracy for each value of k    
# Set the range of k values to consider
accuracy.df <- data.frame(k = seq(1, 20, 1), overallaccuracy = rep(0, 20)) 
for(i in 1:20) 
  {knn.pred <- class::knn(train = train.norm.df, 
                         test = valid.norm.df, 
                         cl = train.df$Personal.Loan, k = i)
  accuracy.df[i, 2] <- confusionMatrix(knn.pred,as.factor(valid.df$Personal.Loan),positive = "1")$overall[1]
}
which(accuracy.df[,2] == max(accuracy.df[,2])) 
plot(accuracy.df$k,accuracy.df$overallaccuracy, main = "Accuracy Vs K", xlab = "k", ylab = "accuracy")

```

### Question - 3
Show the confusion matrix for the validation data that results from using the best k.

#### Confusion Matrix using best K=3
```{r}

knn_result <- class::knn(train = train.norm.df,
                         test = valid.norm.df, 
                         cl = train.df$Personal.Loan, k = 3)

confusionMatrix(knn_result,as.factor(valid.df$Personal.Loan))

```


### Question - 4
Consider the following customer: Age = 40, Experience = 10, Income = 84, Family = 2, CCAvg = 2, Education_1 = 0, Education_2 = 1, Education_3 = 0, Mortgage = 0, Securities Account = 0, CD Account = 0, Online = 1 and Credit Card = 1. Classify the customer using the best k.
```{r}
given_customer2<-data.frame(
  Age = 40, 
  Experience = 10, 
  Income = 84, 
  family =2, 
  CCAvg = 2, 
  Education_1 = 0,
  Education_2 = 1, 
  Education_3 = 0, 
  Mortgage = 0, 
  Securities.Account = 0, 
  CDAccount = 0, 
  Online = 1, 
  CreditCard = 1)

knn_result <- class::knn(train = train.norm.df, 
                       test = given.cust.norm, 
                       cl = train.df$Personal.Loan, k = 3)
knn_result
```
#### Print the predicted customer class
```{r}
if (knn_result == 1) {
  cat("The customer is classified as likely to accept the loan offer (Loan Acceptance = 1).\n")
} else {
  cat("The customer is classified as unlikely to accept the loan offer (Loan Acceptance = 0).\n")
}
```
 
### Question - 5
Repartition the data, this time into training, validation, and test sets (50% : 30% : 20%). Apply the k-NN method with the k chosen above. Compare the confusion matrix of the test set with that of the training and validation sets. Comment on the differences and their reason.

```{r}
# Split the data to 50% training and 30% Validation and 20% Testing 
set.seed(1)
Train_Index1 <- sample(row.names(universal_m.df), 0.5*dim(universal_m.df)[1])
Val_Index1 <- sample(setdiff(row.names(universal_m.df),Train_Index1),0.3*dim(universal_m.df)[1])
Test_Index1 <-setdiff(row.names(universal_m.df),union(Train_Index1,Val_Index1))
Train_Data <- universal_m.df[Train_Index1,]
Validation_Data <- universal_m.df[Val_Index1,]
Test_Data <- universal_m.df[Test_Index1,]

# Normalize the data
train.norm.df1 <- Train_Data[,-10]
valid.norm.df1 <- Validation_Data[,-10]
Test.norm.df1  <-Test_Data[,-10]

norm.values1 <- preProcess(Train_Data[, -10], method=c("center", "scale"))
train.norm.df1 <- predict(norm.values1, Train_Data[,-10])
valid.norm.df1 <- predict(norm.values1, Validation_Data[,-10])
Test.norm.df1 <-predict(norm.values1,Test_Data[,-10])

# Predict using K-NN(k- Nearest neighbors)

validation_knn = class::knn(train = train.norm.df1, 
                           test = valid.norm.df1,  
                           cl = Train_Data$Personal.Loan, 
                           k = 3)

test_knn = class::knn(train = train.norm.df1, 
                     test = Test.norm.df1,  
                     cl = Train_Data$Personal.Loan, 
                     k = 3)

Train_knn = class::knn(train = train.norm.df1, 
                     test = train.norm.df1,  
                     cl = Train_Data$Personal.Loan, 
                     k = 3)
```

### Validation confusion Matrix

```{r}
validation_confusion_matrix = confusionMatrix(validation_knn, 
                                               as.factor(Validation_Data$Personal.Loan), 
                                               positive = "1")

validation_confusion_matrix

```
### Test confusion Matrix

```{r}

test_confusion_matrix = confusionMatrix(test_knn, 
                                         as.factor(Test_Data$Personal.Loan), 
                                         positive = "1")


test_confusion_matrix

```

### Train confusion Matrix

```{r}

Training_confusion_matrix = confusionMatrix(Train_knn, 
                                               as.factor(Train_Data$Personal.Loan), 
                                               positive = "1")

Training_confusion_matrix
```


### Summary of the data and potential reasons for the differences:

Training Set Confusion Matrix: This matrix is based on the data the model was trained on, so it may show high accuracy, potentially close to 100%, especially with k-NN where k = 3. High true positives and true negatives can be expected since the model has seen this data before and has potentially overfit it.

Validation Set Confusion Matrix: This matrix represents the model's performance on new data that it hasn't seen during training.Accuracy may be slightly lower than the training set as the model may not generalize perfectly to new customers.You may observe some false positives and false negatives, indicating that the model may make errors on unseen data.This matrix provides a more realistic estimate of how well the model will perform in practice.

Test Set Confusion Matrix: The test set confusion matrix is the most critical evaluation as it represents completely unseen data, similar to real-world customers.The accuracy may be similar to or slightly worse than the validation set, depending on the dataset's characteristics and the model's generalization ability.Differences between the test set and validation/training sets can be due to variations in data distribution, noise, and the model's adaptability to different patterns.

### Differences in Test vs.Train:

Accuracy: Train has a higher accuracy (0.9772) compared to Test (0.9507).This because of differences in the datasets used for evaluation. Train may have a more balanced or easier-to-predict dataset.

Sensitivity (True Positive Rate): Train has higher sensitivity (0.7589) compared to Test (0.5875). This indicates that Train's model is better at correctly identifying positive cases (e.g., loan acceptances). It may have a lower false negative rate.

Specificity (True Negative Rate): Train has higher specificity (0.9987) compared to Test (0.99403). This suggests that Train's model is better at correctly identifying negative cases (e.g., loan rejections). It may have a lower false positive rate.

Positive Predictive Value (Precision): Train has a higher positive predictive value (0.9827) compared to Test (0.92157). Train's model is more precise in predicting positive cases, resulting in fewer false positive predictions.  


### Differences in Train vs.Validation:  

Accuracy:Train still has a higher accuracy (0.9772) compared to Validation (0.958). Similar to the comparison with Test, Train may have a more balanced or easier-to-predict dataset.

Sensitivity (True Positive Rate): Train has higher sensitivity (0.7589) compared to Validation (0.625). Train's model is better at correctly identifying positive cases. This indicates that Validation's model may have a higher false negative rate.

Specificity (True Negative Rate): Train has higher specificity (0.9987) compared to Validation (0.9934). Train's model is better at correctly identifying negative cases. Validation's model may have a slightly higher false positive rate.

Positive Predictive Value (Precision): Train still has a higher positive predictive value (0.9827) compared to Validation (0.9091). Train's model is more precise in predicting positive cases, resulting in fewer false positive predictions.

### Potential Reasons for Differences:  

Differences in data sets: The composition and distribution of data in various sets can have a big impact on how well a model works. As an example, one data collection could be more unbalanced than another, making it more difficult to forecast unusual events.   

Overfitting: If the training set performs noticeably better than the validation and test sets, overfitting may have occurred. The model might not generalize effectively because it may have trained to suit the training data too closely.

Data Quality: Performance variances may be impacted by variations in data quality, distribution, or the existence of outliers between the sets.

Sample Size: Performance metrics may be more variable if the validation and test sets have smaller sample numbers.

Unpredictability: Variations may also result from the unpredictability of the data segmentation and the particular samples selected for each set.

Tuning of hyperparameters: The choice of k in k-NN or other model-specific parameters, for example, might have an impact on the performance of a model.

